Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Minor tidying

\subsection{The \psykal Approach}

@@

\subsection{The `Shallow' Program}

@@

\begin{figure}
\centering
\includegraphics[width=85mm]{psykal_shallow}
\caption{A schematic of the vanilla \psykal version of the Shallow code.}
\label{FIG_psykal_shallow_structure}
\end{figure}

@@

The primary aim of this work is to determine whether any performance
loss due to the \psykal separation of concerns can be negated by
optimising the PSy layer. To evaluate this we perform tests on a range
of hardware and a range of compilers. The hardware and compilers are
listed in Table~\ref{TABLE_compilers}. Where a compiler is available
on a given piece of hardware, the version number used in this work is
specified.
%
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
'chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
3.3~GHz.

@@

\item Move all kernel subroutines into the same module as as that
  containing the middle/PSy layer (module inlining);

@@

\item Manually in-line (kernel inline) the three field-copy operations
  performed as part of the time-update section;

@@

\item Make the array/loop bounds explicitly constant for the duration
  of the time-stepping loop (pass as arguments to the middle layer
  rather than using module variables).

@@

In addition to these transformations we also applied two
transformations that were suggested by the IBM compiler team. The
first of these manually blocked the outer loop of the time-smoothing
loop nest using a factor of four. The second effectively undoes our 3rd
transformation by fissioning the first loop nest back into three
separate loop nests.

We shall see that these transformations do not always result in improved
performance. Whether or not they do so depends both on the compiler
and the problem size. We also emphasise that the aim of these
optimisations is to recover, as far as is possible, the structure of
the original version of the code. It may well be that transforming the
code into some other structure would result in better performance on a
particular architecture. However, exploring this optimisation space is
beyond the scope of the present work.

@@

% The IBM Fortran
%compiler for Linux on Power 8 is relatively immature and failed to
%SIMD-vectorise two of the three major loop nests in the time-stepping
%loop. Figure~\ref{FIG_orig_perf_summary} therefore also includes the
%performance results obtained after the source had been modified to
%work-around this limitation (we fissioned one loop nest and manually
%unrolled another). Work is currently underway on updating the compiler
%so that these modifications will be unecessary with a future release.

@@

The exceptional case is that of the Gnu-compiled binary on the Haswell
CPU for a problem size of $256^2$.  This is strange since the
performance difference for the same case on the Ivy Bridge CPU is just
0.8\%. {\bf installing 4.9.1 on desktop so can try running this case
  with it. In fact there are considerable difference between
  slow-downs seen with 4.8.2 and 4.9.3 on desktop -things are much
  better with more recent compiler.}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=120mm]{opt_stages_256}
\caption{The performance of the \psykal version of Shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called (`module in-lining') has
the most dramatic effect on the performance of the compiled code; it
now achieves almost 80\% of the performance of the original. The next
most significant transformation is that of manually in-lining the
kernel code itself into the bodies of the loops from which they are
called (`kernel in-lining'). This gets the \psykal version to within
8\% of the performance of the original.  The dramatic performance
improvement obtained from the first transformation demonstrates that
the Gnu compiler is unable to optimise over separate files. Therefore,
the fact that the time-stepping loop in the \psykal version still
retains a call to an I/O routine and a call to the middle-layer, both
of which are in different files, is likely to account for this
remaining performance penalty of 8\%.

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal version of
Shallow achieves some 80\% of the performance of the original
version. Only two code transformations were required to increase this
to 100\%. The first of these was the fusion of the computational
loops. However, we found that performance suffered (purple bar) when
this was extended to include the loop required for field copies at the
end of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
investigation of this slow-down revealed that the compiler was unable
to SIMD vectorise the newly-fused loop after it had in-lined the body
of the kernel called from within it. Once this in-lining was done by
hand (dark-blue bar), the compiler was able to vectorise the loop and
the performance of the original version of Shallow was recovered.

>>>>

\subsection*{The \psykal Approach}

@@

\subsection*{The `Shallow' Program}

@@

\begin{figure}
\centering
\includegraphics[width=100mm]{psykal_shallow}
\caption{A schematic of the vanilla \psykal version of the Shallow code.}
\label{FIG_psykal_shallow_structure}
\end{figure}

@@

The primary aim of this work is to determine whether any performance
loss due to the \psykal separation of concerns can be negated by
optimising the PSy layer. To evaluate this we perform tests on a range
of hardware and a range of compilers. The hardware and compilers are
listed in Table~\ref{TABLE_compilers}. Where a compiler is available
on a given piece of hardware, the version number used in this work is
specified.
%
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB. The CPU in the IBM Power-8 system is built around 12
`chiplets'. Each chiplet has a single core with 64~KB of L1 cache,
512~KB of L2 cache and 8~MB of L3 cache. The L3 cache of each chiplet
is shared with all other chiplets so that there is a total of 96~MB of
last-level cache. The cores on the system we used had a clock speed of
3.3~GHz.

@@

\item Move all kernel subroutines into the same module as as that
  containing the middle/PSy layer (`module inlining');

@@

\item Manually in-line (`kernel inline') the three field-copy operations
  performed as part of the time-update section;

@@

\item Make the array/loop bounds explicitly constant for the duration
  of the time-stepping loop (pass as arguments to the middle layer
  rather than using module variables);

\item Un-roll the outer loop of the time-smoothing loop nest using a
blocking factor of four (IBM supplied);

\item Fission the first loop nest back into three separate loop nests
  (IBM supplied).

@@

As indicated above, the last two of these transformations were
suggested by the IBM compiler team. The second of these effectively
undoes our 3rd transformation by fissioning the first loop nest back
into three separate loop nests.

We shall see that these transformations do not always result in improved
performance. Whether or not they do so depends both on the compiler
and the problem size. We also emphasise that the aim of these
optimisations is typically to recover, as far as is possible, the structure of
the original version of the code. It may well be that transforming the
code into some other structure would result in better performance on a
particular architecture. However, exploring this optimisation space is
beyond the scope of the present work.

@@

@@

The exceptional case is that of the Gnu-compiled binary on the Haswell
CPU for a problem size of $256^2$.  This is strange since the
performance difference for the same case on the Ivy Bridge CPU is just
0.8\%. We investigated this further by compiling and running this case
with versions 4.8.3, 4.9.1 and 4.9.3 of the Gnu compiler on the
Haswell system. These gave slow-downs of 7.8\%, 7.3\% and 6.6\%,
respectively. From these results we conclude that the Gnu compiler's
support for/ability to exploit the newer Haswell architecture is still
maturing.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=123mm]{opt_stages_256}
\caption{The performance of the \psykal version of Shallow for the
  $256^{2}$ domain at each stage of optimisation. Results are given as
  a percentage of the performance of the original code for each
  compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

With the Gnu compiler (first and third clusters in
Figure~\ref{FIG_opt_stages_256}), the vanilla \psykal version of
Shallow only achieves ~5-6\% of the performance of the
original. Simply copying the code for each of the kernel subroutines
into the module from which they are called (module in-lining) has the
most dramatic effect on the performance of the compiled code; it now
achieves more than 80\% of the performance of the original. The
transformations required to get beyond this level then become
architecture dependent: on Haswell loop fusion achieves more than 90\%
while on Ivy Bridge it is fully-fusing the first loop nest. The
dramatic performance improvement obtained from the module-inlining
transformation demonstrates that the Gnu compiler is unable to
optimise over separate files.

With the Intel compiler, the key transformations to recover
performance are different (second and fourth clusters in
Figure~\ref{FIG_opt_stages_256}).  Even the vanilla \psykal version of
Shallow achieves some 80\% of the performance of the original
version. Only two code transformations were required to increase this
to $\sim$100\%. The first of these was the fusion of the computational
loops. However, we found that performance suffered (bar 5) when this
was extended to include the loop required for field copies at the end
of a time-step (Figure~\ref{FIG_time_smooth_code}). Further
investigation of this slow-down revealed that the compiler was unable
to SIMD vectorise the newly-fused loop after it had in-lined the body
of the kernel called from within it. Once this in-lining was done by
hand (bar 6), the compiler was able to vectorise the loop and the
performance of the original version of Shallow was recovered on the
Haswell system. On Ivy Bridge the first loop nest also had to be fully
fused (bar 7) to recover $\sim100\%$ of the performance.
